---
title: "Decile Analysis"
author: "VijayUpadhyay"
date: "10 October 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction 

We have a data set of >22000 customers on retail stores, we want to analyse the data based on total sales per customer or total buying amount of customers to find the most valuable customers.

Following is the description of parameters in dataset:

|Parameters             | Description                                                                         |
|-----------------------|-------------------------------------------------------------------------------------|
| cust_id               | customer Identity                                                                   |
| No_of_days_frm_L_pur  | number of days from last purchase                                                   |
| total_sale            | total buy/sale value                                                                |
| No_baskets            | number of times customer purchased/checked-out from a store or number of billings   |
| item_cnt              | number of items bought                                                              |
| no_SKUs               | number of stock keeping units                                                       |
| no_Categories         | number of categories from which items purchased                                     |
| avg_no_prds_w_ret     | average number of products per trip                                                 |
| Flag_Multi_cat_p2     |                                                                                     |
|                       |                                                                                     |

### Import and Analyze data file
```{r}

dir <- "/home/vijayupadhyay/Desktop/AnalytixLabs/4-BA/class5/RFMSegmentation"
setwd(dir)
customerData <- read.csv("data/dataRFMValueBasedSegmentation.csv", header = TRUE)

#names(customerData)
str(customerData, list.len = 150)

```

### Creating Deciles:

In the following section we look at the bacis summary stats of "total_sale" and add a "decile" column to the dataset specifying the decile group customer belongs to based on "total_sale".

```{r}

# Basic statistics about the sales data
minSale <- min(customerData$total_sale, na.rm = TRUE)
maxSale <- max(customerData$total_sale, na.rm = TRUE)
sumSale <- sum(customerData$total_sale, na.rm = TRUE)
meanSale <- mean(customerData$total_sale, na.rm = TRUE)
medianSale <- median(customerData$total_sale, na.rm = T)
sdSale <- sd(customerData$total_sale, na.rm = T)

summarySales <- data.frame(minSale, meanSale, medianSale, sdSale, maxSale, sumSale)
summarySales

# Find the decile locations based on total sale
decLocations <- quantile(customerData$total_sale, probs = seq(0.1,0.9,by=0.1))

# Add decile value to dataset
customerData$decile <- findInterval(customerData$total_sale,c(-Inf,decLocations, Inf))

#Checking the sizes of Deciles
#table(customerData$decile)
xtabs(~decile,customerData)

```

### Summarizing Parameters

In this section we create a final table which summarizes all the parameters for every decile group of "total_sales" and store it on our current working directory.

```{r,message=FALSE}

library(dplyr)
options(dplyr.width = Inf)

#Calculating Totals
totals<-summarise(customerData, sum_sales = sum(total_sale, na.rm=T),
                  sum_baskets = sum(No_baskets, na.rm=T), sum_item_cnt = sum(item_cnt, na.rm=T))

byDecile <- group_by(customerData, decile)

summaryDecile <- summarise(byDecile, totalSales = sum(total_sale, na.rm=T),
                           pctTotalSales = sum(total_sale, na.rm=T)/totals$sum_sales,
                           avgSales = mean(total_sale, na.rm=T),
                           avgBasketSize = sum(total_sale, na.rm=T)/sum(No_baskets, na.rm=T),
                           pctBaskets = sum(No_baskets, na.rm=T)/totals$sum_baskets,
                           itemsPerBasket = sum(item_cnt, na.rm=T)/sum(No_baskets, na.rm=T),
                           numCategoriesPerBasket = sum(no_Categories, na.rm=T)/sum(No_baskets, na.rm=T),
                           pricePerItem = sum(total_sale, na.rm=T)/sum(item_cnt, na.rm=T),
                           pctMultiCategoryBuyers = mean(Flag_Multi_cat_p2, na.rm=T)
)

summaryOverall <- summarise(customerData, totalSales = sum(total_sale, na.rm=T),
                            pctTotalSales = sum(total_sale, na.rm=T)/totals$sum_sales,
                            avgSales = mean(total_sale, na.rm=T),
                            avgBasketSize = sum(total_sale, na.rm=T)/sum(No_baskets, na.rm=T),
                            pctBaskets = sum(No_baskets, na.rm=T)/totals$sum_baskets,
                            itemsPerBasket = sum(item_cnt, na.rm=T)/sum(No_baskets, na.rm=T),
                            numCategoriesPerBasket = sum(no_Categories, na.rm=T)/sum(No_baskets, na.rm=T),
                            pricePerItem = sum(total_sale, na.rm=T)/sum(item_cnt, na.rm=T),
                            pctMultiCategoryBuyers = mean(Flag_Multi_cat_p2, na.rm=T)
)

decileProfiling <- rbind(summaryDecile, cbind(decile="Overall", summaryOverall))
write.csv(decileProfiling, "decileProfiling.csv")
decileProfiling

```
