---
title: "RFM Analysis"
author: "VijayUpadhyay"
date: "10 October 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction:

While planning for marketing spend, or formulating a new promotion, retail marketers employ sophesticated techniques to segment and target customers. Because it would be a waste of marketing spend if, for example, an ad campaign is targeted to all the customers who ever bought from you. Such an untargeted marketing promotion is unlikely to have a high conversion rate and may even hurt your brand value.

RFM analysis is a useful method to improve customer segmentation by dividing customers into various groups for future personalization services and to identify customers who are more likely to respond to promotions. Therefore it can help retailers maximize the return on their marketing investments. Under RFM analysis, each customer is scored based on three factors, namely Recency, Frequency, and Monetary value. RFM analysis can help companies identify customers that are most likely to respond to a new offer. Following is a detailed explaination of its factors:

**Recency:** Recency is the most important predictor of who is more likely to respond to an offer. Customers who have purchased recently are more likely to purchase again compared to those who did not purchase recently.

**Frequency:** The second most important factor is how frequently these customers purchase. The higher the frequency, the higher is the chances of them responding to offers.

**Monetary:** The third factor is the amount of money these customers have spent on purchases. Customers who have spent higher are more likely to purchase based on the offer compared to those who have spent less.

We have a data set of >22000 customers on retail stores, we want to analyse the data using RFM classification of customers.
Following is the description of parameters in dataset:

|Parameters             | Description                                                                         |
|-----------------------|-------------------------------------------------------------------------------------|
| cust_id               | customer Identity                                                                   |
| No_of_days_frm_L_pur  | number of days from last purchase                                                   |
| total_sale            | total buy/sale value                                                                |
| No_baskets            | number of times customer purchased/checked-out from a store or number of billings   |
| item_cnt              | number of items bought                                                              |
| no_SKUs               | number of stock keeping units                                                       |
| no_Categories         | number of categories from which items purchased                                     |
| avg_no_prds_w_ret     | average number of products per trip                                                 |
| Flag_Multi_cat_p2     |                                                                                     |
|                       |                                                                                     |

### Import and Analyze data file

```{r}

dir <- "/home/vijayupadhyay/Desktop/AnalytixLabs/4-BA/class5/RFMSegmentation"
setwd(dir)
customerData <- read.csv("data/dataRFMValueBasedSegmentation.csv")

str(customerData)

```


### Classifying the Data Set

To perform RFM analysis, each customer is assigned a score for recency, frequency, and monetary value, and then a final RFM score is calculated.

Recency score is calculated based on the date of their most recent purchase. The scores are generally categorized based on the values. For example, a company may follow a category system of 1 to 5, highest to lowest value could be 1 to 5 or 5 to 1. In our case we are using quartile locations in recency data distribution which is represented by **No_of_days_frm_L_pur**(number of days since last puchase) and assigning 1 to most recent customers and 4 to customers with oldest purchase date.

Similarly, frequency score is calculated based on the number of times the customers purchased. Customers with higher frequency receive a higher score. In our dataset frequency is represented by **No_baskets**. We use quartile locations in its distribution to assign classes as we see in the distribution most of the customers have number of baskets between 1-5 and thus we are getting 3 almost equal partition of the data. So the levels here are 1 to 3, 1 being least frequent and 3 being most frequent.

Finally, customers are assigned a score based on the amount they spent on their purchases. For calculating this score, we use **total_sale** parameter. Again we use quartile location points to classify the parameter and assign them values 1 to 4 =. Where 1 represents least spending customers and 4 represent most spending customers.


```{r}

#recency: number of days since last purchase
recencyLocation <- quantile(customerData$No_of_days_frm_L_pur, probs = seq(0.25, 0.75, 0.25))
#frequency: number of baskets
frequencyLocation <- quantile(customerData$No_baskets, probs = seq(from = 0.25, to = 0.75,by = 0.25))
#monetory: total sale
monetaryLocation <- quantile(customerData$total_sale, probs = seq(0.25,0.75, 0.25))

#creat columns

customerData$recency <- findInterval(customerData$No_of_days_frm_L_pur, c(-Inf, recencyLocation, Inf))
customerData$frequency <- findInterval(customerData$No_baskets, c(frequencyLocation, Inf))
customerData$monetary <- findInterval(customerData$total_sale, c(-Inf, monetaryLocation, Inf))
```

### Visualizing Partitions
```{r}

recencyLocation
xtabs(~recency, customerData)

frequencyLocation
xtabs(~frequency, customerData)

monetaryLocation
xtabs(~monetary, customerData)

segSize <- ftable(xtabs(~recency+frequency+monetary,customerData))
prop.table(segSize)*100

segVal <- xtabs(total_sale~frequency+monetary,customerData)
prop.table(segVal)*100

```


### Assigning Lables to Classes and Final Representation


In this section what we would like to achive is to assign different categories to different customers based on the following table. 



| R | F |          |         M|          |          |
|:-:|:-:|:--------:|:--------:|:--------:|:--------:|
|   |   |     1    |     2    |     3    |     4    |
|   | 1 | Standard | Standard |  Silver  |  Silver  |
| 1 | 2 | Standard |  Silver  |  Silver  |   Gold   |
|   | 3 |  Silver  |  Silver  |   Gold   |  Premium |
|   |   |          |          |          |          |
|   | 1 | Standard | Standard |  Silver  |  Silver  |
| 2 | 2 | Standard |  Silver  |  Silver  |   Gold   |
|   | 3 |  Silver  |  Silver  |   Gold   |  Premium |
|   |   |          |          |          |          |
|   | 1 | Standard | Standard |  Silver  |  Silver  |
| 3 | 2 | Standard |  Silver  |  Silver  |   Gold   |
|   | 3 |  Silver  |  Silver  |   Gold   |  Premium |
|   |   |          |          |          |          |
|   | 1 | Standard | Standard |  Silver  |  Silver  |
| 4 | 2 | Standard |  Silver  |  Silver  |   Gold   |
|   | 3 |  Silver  |  Silver  |   Gold   |  Premium |
|   |   |          |          |          |          |



In addition Customers **recency equals to 1** i.e. most recent users will be labeled as **Active**, then customers with **recency equals to 2 or 3** will be labeled as **At Risk** and the last group of customers where **recency equals to 4** will be labeled as **Churn**.
Then finally we produce some output tables to analyse the customer data.


```{r}

customerData$recencyLable[customerData$recency == 1] <- "Active"
customerData$recencyLable[customerData$recency == 2 | customerData$recency == 3] <- "At Risk"
customerData$recencyLable[customerData$recency == 4] <- "Churn"

customerData$classLable[customerData$frequency + customerData$monetary == 7] <- "Premium"
customerData$classLable[customerData$frequency + customerData$monetary == 6 ] <- "Gold"
customerData$classLable[(customerData$frequency+customerData$monetary == 4)|(customerData$frequency+customerData$monetary == 5)] <- "Silver"
customerData$classLable[is.na(customerData$classLable)] <- "Standard"

customerData$recencyLable <- factor(customerData$recencyLable)
customerData$classLable <- factor(customerData$classLable, levels = c("Premium", "Gold", "Silver", "Standard"))

customersEachClass <- xtabs(~classLable+recencyLable,customerData)
salesByClasss <- xtabs(total_sale~classLable+recencyLable,customerData)

prop.table(salesByClasss)*100

```